---
title: 'R: Youtube Summary Notes'
date: 'Last Updated: May 14, 2023'
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
    always_allow_html: yes
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```
# Purpose
This file contains links and notes to Youtube videos describing R functions and concepts. Datasets used in this document are either preloaded in R, or are custom and posted below each header. It is recommended to load/install tidyverse as pipes (%>%) are frequently used, and many packages covered in this document are part of tidyverse. 
```{r}
#load tidyverse
library(tidyverse)
```

# Visualizations

## Simple Tables
Package: Base R, Dataset: iris

### Frequency Table:  A frequency table shows the count for one variable
Video Link: [3:13](https://www.youtube.com/watch?v=9gJJi_HrnCA&t=3m13s)
```{r}
table(iris$Species)
```

### A proportion table shows the proportion for one variable
Video Link: [7:15](https://www.youtube.com/watch?v=9gJJi_HrnCA&t=7m15s)
```{r}
prop.table(table(iris$Species)) 
```

### Table with percentages 
Video Link: [5:45](https://www.youtube.com/watch?v=9gJJi_HrnCA&t=5m45s)
```{r}
#multiply prop.table formula by 100
prop.table(table(iris$Species))*100 
```
### A contingency table shows the count for two or more variables 
Video Link: [8:32](https://www.youtube.com/watch?v=9gJJi_HrnCA&t=8m32s)
```{r}
#The petal width count for 3 different flower species are shown. 
xtabs(~Petal.Width +Species, iris) 
```

## Presentation Summary Tables
Package: gtsummary, Dataset: CO2

### Install/load gtsummary 
Video Link: [0:10](https://www.youtube.com/watch?v=gohF7pp2XCg&t=10s)
```{r}
library(gtsummary)
```

### Basic Summary Table
Video Link: [0:44](https://www.youtube.com/watch?v=gohF7pp2XCg&t=44s)
```{r}
CO2 %>% select(!c(Plant,conc)) %>% #Removing ‘Plant’ and ‘conc’ variables
  tbl_summary()   #creates summary table 
```

### Summary split by categorical variable 
Video Link: [1:42](https://www.youtube.com/watch?v=gohF7pp2XCg&t=1m42s)
```{r}
CO2 %>% select(!c(Plant,conc)) %>% 
  tbl_summary(by = Type) #organizes the data by the categorical variable 'Type'
```

### Summary split by categorical variable with p-values 
Video Link: [2:13](https://www.youtube.com/watch?v=gohF7pp2XCg&t=2m13s)
```{r}
CO2 %>% select(!c(Plant,conc)) %>% 
  tbl_summary(by = Type) %>% 
  add_p() # adds p-values to the table 
```

### Summary including overall, extra heading, and other statistics 
Video Link: [3:02](https://www.youtube.com/watch?v=gohF7pp2XCg&t=3m02s)
```{r}
CO2 %>% select(!c(Plant,conc)) %>% 
  tbl_summary(by = Type, 
              statistic = list(all_continuous() ~ "{mean} ({sd})", #adds mean and standard deviation
              all_categorical() ~ "{n} / {N} ({p}%)" ), #add percentage 
              digits = all_continuous() ~ 2) %>% #rounds decimal to 2 digits
  add_overall() %>% #adds total column 
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Location**") #adds title 'Location'
```
### Create crosstab with p-values 
Video Link: [4:27](https://www.youtube.com/watch?v=gohF7pp2XCg&t=4m27s)
```{r}
#Crosstab shows the relationship between two categorical variables 

CO2 %>% #dataset
  tbl_cross(row = Type, col = Treatment, percent = "cell") %>%  #creates cross tables w/ Type and Treatment 
    add_p() #adds p-values
```

## Barplot (1)
Package: Base R, Dataset: custom 

### Load Custom data 
Video Link: [0:10](https://www.youtube.com/watch?v=pYbuWU77QkU&t=0m10s)
```{r}
values <- c(.4, .75, 0.2, 0.6, 0.5) 
```

### Make simple barplot
Video Link: [0:30](https://www.youtube.com/watch?v=pYbuWU77QkU&t=0m30s)
```{r}
barplot (values) 
```

### Add color to barplot 
Video Link: [1:23](https://www.youtube.com/watch?v=pYbuWU77QkU&t=1m23s)
```{r}
#Add color by using the text characters  
barplot(values, col = "#1b98e0") 
        
#Add color by using color name
barplot(values, col = "darkgreen")
```
### Change bar orientation to horizontal 
Video Link: [2:00](https://www.youtube.com/watch?v=pYbuWU77QkU&t=2m00s)
```{r}
barplot(values, horiz = TRUE)
```

### Add labels to barplot 
Video Link: [2:27](https://www.youtube.com/watch?v=pYbuWU77QkU&t=2m27s)
```{r}
#Create a vector called group that contains that labels for the barplot 
group <- LETTERS[1:5]

#Assign the new vector to ‘names.arg’
barplot(values, names.arg = group)
```

### Creating a stacked bar plot
Video Link: [3:33](https://www.youtube.com/watch?v=pYbuWU77QkU&t=3m33s)
```{r}
# Create matrix called data for the data that includes two different values for each column/label
 data <- as.matrix(data.frame(  #as.matrix converts the dataframe to a matrix
   A = c(0.2, 0.4), 
   B = c(0.3,0.1), 
   C = c(0.7, 0.1), 
   D = c(0.1, 0.2),  
   E = c(0.3, 0.3)))
# Create row names for the matrix called Group 1 and Group 2 
rownames(data) <- c ("Group 1", "Group 2")

#Plot stacked bar plot with matrix data with 2 different colors 
barplot(data, col = c("#1b98e0", "#353436"))
```

### Add legend to bar plot 
Video Link: [5:04](https://www.youtube.com/watch?v=pYbuWU77QkU&t=5m04s)
```{r}
#legend must be executed with barplot in the same code chunk
barplot(data, col = c("#1b98e0", "#353436"))

legend("topright",  #location of legend
       legend = c("Group 1", "Group 2"), #names of legend
       fill = c("#1b98e0", "#353436"))  #fill colors for legend 
```

### Grouped barchart (columns side-by-side) 
Video Link: [6:25](https://www.youtube.com/watch?v=pYbuWU77QkU&t=6m25s)
```{r}
#The groups should be some type of categorical data 
barplot(data, 
        col = c ("#1b98e0", "#353436"),
        beside = TRUE) #places columns beside each other 
```

### Manually grouped barchart: Color subgroups by mean then create grouped barplot
Video Link: [24:28](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=24m28s)
```{r}
#Using diamonds pre-loaded dataset

#Create object containing mean price by subgroup clarity 
diamonds_m_cl_co <- aggregate(diamonds, price ~ clarity + color, mean) 

#Plot grouped barplot with object containing subgroups 
ggplot(diamonds_m_cl_co, aes(x=clarity, y = price, fill = color)) + 
  geom_bar(stat = "identity",  position = "dodge") #dodge makes barplot grouped. 
```

## Barplot (2)

### Custom Data and load ggplot2
Video Link: [7:48](https://www.youtube.com/watch?v=pYbuWU77QkU&t=7m48s)
```{r}
#Custom data 
values <- c(.4, .75, 0.2, 0.6, 0.5) 

#load ggplot2
library(ggplot2) #automatically loaded with tidyverse 
```

### Ggplot2 only takes dataframes as input, not matrices like baseR 
Video Link: [8:09](https://www.youtube.com/watch?v=pYbuWU77QkU&t=8m09s)
```{r}
#Dataframe is a slightly different way to store data than a matrix. 
data_ggp <- data.frame(group, values)

#If unsure of your data type, run ‘typeof(name_of_data)’
typeof(data_ggp) #should now say dataframe

```

### Create a bar plot in ggplot2
Video Link: [8:47](https://www.youtube.com/watch?v=pYbuWU77QkU&t=8m47s)
```{r}
ggplot(data_ggp, aes(x = group, y = values)) + 
  geom_bar(stat = "identity") #specifies a bar chart based on our data
```

## Ordering Bars of a Barplot 

### Custom data 
Video Link: [0:15](https://www.youtube.com/watch?v=mpHaEP21ZQE&t=0m15s)
```{r}
data <- data.frame(
  x = c("A", "B", "C", "D", "E"),  
  y = c(0.5, 2, 1.2, -0.2, 0.7))
```

### Manually ordering bars 
Video Link: [1:55](https://www.youtube.com/watch?v=mpHaEP21ZQE&t=1m55s) 
```{r}
# Create duplicate of data
data1 <- data 

# Change factor levels of data

#changes the factor ordering to ‘B’, ‘D’, ‘E’, ‘C’, ‘A’
data1$x <- factor(data1$x, levels = c("B", "D", "E", "C", "A")) 

#Default factor ordering is ‘A’, ‘B’, ‘C’, ‘D’, ‘E’
```

### Order bars in increasing order 
Video Link: [3:30](https://www.youtube.com/watch?v=mpHaEP21ZQE&t=3m30s)
```{r}
# Step 1: Duplicate  data 
data2 <- data 

#Step 2: Change factor levels to increasing order ()

#Replace x and y with variable names used in your plot
data2$x <- factor(data2$x, levels = data2$x[order(data2$y)])

#Step 3: Create plot with new data 
 ggplot(data2, aes(x, y)) + 
   geom_bar(stat = "identity")
```

### Order bars in decreasing order 
Video Link: [4:30](https://www.youtube.com/watch?v=mpHaEP21ZQE&t=4m30s)

```{r}
#Step 1: Duplicate data 
data3 <- data         

#Step 2: Change factor levels to decreasing order 
 data3$x <- factor(data3$x, levels = data3$x[order(data3$y, decreasing = TRUE)])
#Replace ‘TRUE’ with ‘FALSE’ to organize in increasing order

#Step 3: Create plot with new data 
ggplot(data3, aes(x, y)) + 
  geom_bar(stat = "identity")
```

## Scatterplots, Changing X-Axis Range, and Facet Layers 
Package: ggplot2, Dataset: custom 

### Custom data 
Video Link:[2:22](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=2m22s)
```{r}
data <- data.frame(
  x = 1:9, 
  y = c(3, 1, 4, 3, 5, 2, 1, 2, 3), 
  group = rep(LETTERS[1:3], each = 3))  
```

### Create base layer for a variety of plots 
Video Link: [3:22](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=3m22s)
```{r}
#shows just x and y axes, and tick marks
ggplot(data, aes(x = x, y = y)) 
```

### Create scatter plot 
Video Link: [3:55](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=3m55s)
```{r}
ggplot(data, aes(x = x, y = y)) + 
  geom_point() #specifies  a scatter plot

```

### Scatter plot: Change point size 
Video Link: [4:45](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=4m45s)
```{r}
ggplot(data, aes(x = x, y = y)) + 
  geom_point(size =3) #changes point size 
```

### Scatter plot: Specify the colors by group 
Video Link: [5:10](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=5m10s)
```{r}
ggplot(data, aes(x = x, y = y, col = group)) + #col = group specifies the color grouping
  geom_point(size =3) + 
  theme(legend.position = "none") #removes default legend  
```

### To save a plot, assign an object to ggplot2 code 
Video Link: [6:08](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=6m08s)
```{r}
ggp_simple <- ggplot(data, aes(x = x, y = y, col = group)) +
  geom_point(size =3 ) #this save the plot to ggp_simple

ggp_simple #the plot won't appear unless the object 'ggp_simple' is called 
```

### Change the x-axis range 
Video Link: [7:01](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=7m01s)
```{r}
ggp_simple + 
  scale_x_continuous(limits = c(-3, 15)) #changes the numeric range of x-axis 
```

### Manually change colors of points by groups 
Video Link: [7:52](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=7m52s)
```{r}
ggp_simple + 
  scale_color_manual(
    breaks = c("A", "B", "C"),  
    values = c("#1b98e0", #selects three different colors. One for each group A, B, C. 
               "#353436", 
               "#e32f08"))
```

### Add multiple scale layers by adding a ‘+’ between layers 
Video Link: [8:58](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=8m58s)
```{r}
ggp_simple +  
  scale_x_continuous(limits = c(-3, 15)) + # the '+' allows multiple layers to be added
  scale_color_manual(
    breaks = c("A", "B", "C"),  
    values = c("#1b98e0", 
               "#353436", 
               "#e32f08"))
```

### Facet layers: Creates side-by-side subplots of a variable or dataset
Video Link: [9:45](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=9m45s)
```{r}
#Facet layers are helpful when your original plot has too much data 
ggp_simple +  
  scale_x_continuous(limits = c(-3, 15)) + 
  scale_color_manual(
    breaks = c("A", "B", "C"),  
    values = c("#1b98e0", 
               "#353436", 
               "#e32f08")) + 
    facet_wrap(group ~ .) + #adds facet layer
    theme_bw() # changes  graph background from gray to white 
```

## Density Plots 
Package: ggplot2, Dataset: diamonds

### Add regression line to facet wrapped subplots 
Video Link: [18:04](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=18m04s) 
```{r}
ggplot(diamonds, aes(x = price, y = carat)) +
  geom_point () + 
facet_wrap(clarity ~ . ) + 
  geom_smooth(method = "lm", formula = y ~ x) #adds regression line
```

### Density plot: Helpful plot for showing distribution of a number
Video Link: [19:44](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=19m44s) 
```{r}
#For density plots only input one column/variable since density will be graphed on the y-axis
ggplot(diamonds, aes(x = depth)) + #variable is depth
  geom_density()
 
```

### Density plot: Draw density plots by group 
Video Link: [20:36](https://www.youtube.com/watch?v=Cl9yE_PFV-0&t=20m36s)
```{r}
#‘fill’ colors the area between the density line and the x-axis, as opposed to coloring just the line 
ggplot(diamonds, aes(x = depth, fill =cut)) + # adds multiple density graphs to plot 
  geom_density(alpha = .3) #makes color more translucent. Can be set to any number lower than 1
```

## Pie Charts 
Package: ggplot2, Dataset: diamonds 

### Step 1: Assign colors to object 
Video Link: [0:33](https://www.youtube.com/watch?v=bzRD_nvvIVI&t=0m33s)

```{r}
#create vector called 'colors' that has 7 different colors
colors <- c("#FFFFFF","#F5FCC2","#E0ED87","#CCDE57", 
            "#B3C732","#94A813","#718200")
```

### Step 2: Shape dataset to have 3 columns: categorical variable, count, and percentage   
Video Link: [0:37](https://www.youtube.com/watch?v=bzRD_nvvIVI&t=0m37s)
```{r}
data <- diamonds %>%   
  group_by(color) %>%   #categorical column 
  summarize(
    counts = n(),      #count column
    percentage = n()/nrow(diamonds))  #percentage column
```

### Step 3: Create Pie Chart 
Video Link: [0:56](https://www.youtube.com/watch?v=bzRD_nvvIVI&t=0m56s)
```{r}
pie <- ggplot(data = data, aes(x="", y = percentage, fill = color)) + 
  geom_col(color = "black") +  #outline color of pie chart 
  coord_polar("y", start = 0) +  #creates pie chart. Following code is formatting.
  geom_text(aes(label = paste0(round(percentage*100), "%")), #adds % lables to pie slices
            position = position_stack(vjust = 0.5)) + #positions labels
  theme(panel.background = element_blank(),  #removes default background
        axis.line = element_blank(),        #removes axis line   
        axis.text= element_blank(),         #removes axis text
        axis.ticks = element_blank(),       #removes axis ticks
        axis.title=element_blank(),         #removes title
        plot.title = element_text(hjust = 0.5, size = 18)) + #format for title
  ggtitle("Pie chart of Diamond Color") + #title
  scale_fill_manual(values = colors) 

pie #call pie chart 


```

## Bubble Plots 
Package: ggplot2, Datasets: mtcars 

Note: A bubble plot is a scatter plot except it has a 3rd numeric variable mapped to a size aesthetic 
Video Link: [0:05](https://www.youtube.com/watch?v=ZMmZPznXtmw&t=5s)

### Subset mtcars dataset 
Video Link: [0:25](https://www.youtube.com/watch?v=ZMmZPznXtmw&t=25s)
```{r}
data <- mtcars %>% mutate(
  cyl = factor(cyl), #creates factor levels 4,6, and 8 for 'cyl' var. Needed for color grouping for code chunks below
  Model = rownames(mtcars)) #creates 'Model' var that contains all the names of cars
```

### Create basic bubble plot
Video Link: [0:30](https://www.youtube.com/watch?v=ZMmZPznXtmw&t=30s)
```{r}
plot1 <- data %>% 
  ggplot(aes(x = wt, y = mpg, size = hp)) + #‘size = hp’ is the 3rd variable that makes this a bubble plot 
  geom_point(alpha = 0.5) #'alpha .5' makes bubbles more transparent
  
plot1
```

### Add color and custom bubble size to bubble plot 
Video Link: [1:09](https://www.youtube.com/watch?v=ZMmZPznXtmw&t=1m09s)
```{r}
#‘color = cyl’ adds color by grouping
plot2 <- data %>% 
  ggplot(aes(x = wt, y = mpg, size = hp, color = cyl, label = Model)) + 
  geom_point(alpha = 0.5) + 
  scale_size(range = c(.1, 15))

plot2

```

### Convert ggplot bubble plot into plotly plot 
Video Link: [1:40](https://www.youtube.com/watch?v=ZMmZPznXtmw&t=1m40s)

```{r}
#Plotly maps are interactive. You can filter data by clicking on the legend and obtain point-specific data by hovering the cursor over a bubble point .

#Step 1: Load/install plotly
library(plotly)

#Step 2: Convert ggplot to plotly plot
p <- ggplotly(plot2, width=500, height=500) %>%  #converts ggplot to plotly
  layout(xaxis = list(range = c(1, 6)),  #formatting
         yaxis = list(range = c(8, 35)), 
         legend = list(x = 0.825, y = .975))

p


```



# Data Management 


## Using the Which and Order Command 
Package: Base R, Dataset: iris 


### Find a row’s values at a specified variable’s maximum value 
Video Link: [0:23](https://www.youtube.com/watch?v=twx3ocWtj2k&t=23s)
```{r}
#Returns the entire row where the variable ‘Sepal’ is the highest value
iris[which.max(iris$Sepal.Length), ] #Be sure to add a comma before the last bracket
```

### Find a row’s values at a specified variable’s minimum value 
Video Link: [1:22](https://www.youtube.com/watch?v=twx3ocWtj2k&t=1m22s)
```{r}
#Returns the entire row where the variable ‘Sepal’ is the lowest value
iris[which.min(iris$Sepal.Length), ]
```


### ‘order’ selects any location of a variable (i.e. 10th or 23rd highest value) 
Video Link: [1:40](https://www.youtube.com/watch?v=twx3ocWtj2k&t=1m40s)
```{r}
#Returns the entire row where the variable ‘Sepal’ is the 11th highest
iris[order(iris$Sepal.Length) [11], ]
```

## Using File Paths 
(Package: Base R)

### Best way use to specify file/directory path is with function ‘file.path’
Video Link: [0:38](https://www.youtube.com/watch?v=M72DjKNPVDA&t=38s) 
```{r}
#change the words in quotation marks to a directory path on your computer 
my_directory <- file.path("C:", "Users", "Joach", "Desktop")

#my_directory can  be called or used in code to refer to the directory path
```

### Function file.path can also be used to specify a specific file 
Video Link: [1:44](https://www.youtube.com/watch?v=M72DjKNPVDA&t=1m44s)
```{r}
#If ‘my_file.csv’ wasn’t added, this would specify a directory path
My_file <- file.path("C:", "Users", "Joach", "Desktop", "my_file.csv")

```

 

## Handeling NAs in R 
Package: Base R, Dataset: airquality 


### Find missing values (NA) 
Video Link: [1:05](https://www.youtube.com/watch?v=q8eR2suCyGk&t=1m05s)
```{r}
#Returns matrix where TRUE is a missing value, FALSE is a numeric value
is.na(airquality)
```

### Count the amount of missing values in a data set 
Video Link: [2:00](https://www.youtube.com/watch?v=q8eR2suCyGk&t=2m00s)
```{r}
#Returns the total amount of missing values 
sum(is.na(airquality))
```

### Remove all missing values from a dataset 
Video Link: [3:10](https://www.youtube.com/watch?v=q8eR2suCyGk&t=3m10s)
```{r}
#Deletes entire row where the missing value is present.
na.omit(airquality)
```

### Calculate the mean of a variable where there is missing values 
Video Link: [4:15]() 
```{r}
mean(airquality$Ozone, na.rm = TRUE) 
```

# Data Analysis

## Using lapply and sapply 
Package: Base R, Dataset: mtcars

Note 1: If the data input is a column, sapply and lapply will apply the function to each value in the specified column. 

Note 2: If the data input is a dataset, sapply and lapply will apply the function to each column.

Note 3: Using these functions is often quicker than using a for loop. 

### lapply runs a function on each element of a dataset and returns a list 
Video Link: [0:38](https://www.youtube.com/watch?v=M72DjKNPVDA&t=38s)
```{r}
#Step 1: Load data (0:45)
data <- mtcars 

#Step 2: Create demo function (0:50)
mpg_category <- function(mpg){ #creates function called mpg_categrory
  if(mpg > 30){   #if mpg is greater than 30, "High" will be assigned
    return("High")
  } 
  else if (mpg > 20){ #if mpg is less than 30 but greater than 20, "Medium"
      return("Medium")
  } else if(mpg <21){
    return("Low")   #If not assigned "Medium" or "High", assign "Low" 
  }
  } 

#Step 3: Use lapply (1:10)

#applies function created above to every value of data$mpg and returns a list. 
lapply(X = data$mpg, FUN = mpg_category)
```

### sapply runs a function on each element of a dataset and returns a vector or matrix 
Video Link: [1:57](https://www.youtube.com/watch?v=34sbvhr_pm8&t=1m57s)
```{r}
#applies function to each value of mpg
sapply(X = data$mpg, FUN = mpg_category) #returns a vector
```



